Manuscript

### 2019-12-16 manuscript.Rmd
update references and methods in manuscript

Cancer cell lines are a long-standing model for systematic testing of candidate therapeutics, beginning with the National Cancer Institute 60 (NCI60) assay from the late 1980s [@alley1988; @shoemaker2006; @stinson1992], which has been used to screen over 100,000 compounds as of 2010 [@holbeck2010]. Since then, numerous small-molecule and gene essentiality screens of various scales and study aims have been conducted in cell lines, from grouping drugs by therapeutic target similarity [@greshock2010] to screening only near-haploid cell lines to generate genome-level insights into gene essentiality [@wang2015a] to broadly identifying cancer dependencies with large-scale screens in multiple cancer types [@mcdonald2017; @mcfarland2018; @meyers2017; @patel2017] or select lineages [@heiser2009; @marcotte2012; @patel2017]. Despite their widespread use, cancer cell lines are known to have issues with inconsistent naming conventions and contamination [@yu2015], and some cell lines have been shown to vary widely at the genetic level and in response to drug treatment across strains [@ben-david2018]. Additionally, comparisons between cancer cell lines and tumors indicate that cell lines have higher numbers of genomic aberrations [@domcke2013; @mouradov2014; @neve2006] and tend to be hypermethylated [@paz2003; @varley2013], which could prove an impediment to translating cell line discoveries into the clinic.

There have also been debates over consistency between drug screen datasets, namely the Broad Institute’s and Novartis Institutes for Biomedical Research’s Cancer Cell Line Encyclopedia (CCLE) [@barretina2012; @consortium2015] and the Genomics of Drug Sensitivity in Cancer (GDSC) from the Cancer Genome Project at the Wellcome Sanger Institute and the Center for Molecular Therapeutics at Massachusetts General Hospital Cancer Center [@garnett2012; @yang2012]. The GDSC has also been referred to as the Cancer Genome Project (CGP) and the Sanger dataset. Studies have shown that drug-gene interactions matched between CCLE and GDSC exhibited poor correlation and inconsistencies [@haibe-kains2013; @jang2014], prompting other groups to join the debate on how best correct for experimental and methodological variation between the original drug screens and subsequent computational analysis [@consortium2015; @geeleher2016; @geeleher2016a; @hatzis2014; @haverty2016; @safikhani2016; @safikhani2017].

@@

To curate an evidence-based list of these clinical biomarkers, we utilized a new oncological data aggregation project conducted by the Variant Interpretation for Cancer Consortium (VICC; https://cancervariants.org/), a Driver Project for the Global Alliance for Genomics and Health (GA4GH) [@theglobalallianceforgenomicsandhealth2016]. The VICC has curated annotations of known cancer variants at varying levels of evidence from multiple resources [@wagner2018]: the Precision Medicine Knowledgebase [@huang2016], MolecularMatch (https://www.molecularmatch.com/index.html), OncoKB [@chakravarty2017], Jackson Labs Clinical Knowledgebase [@patterson2016], Clinical Interpretations of Variance in Cancers (CIViC) [@griffith2017], and the Cancer Genome Interpreter (CGI) [@tamborero2018]. These harmonized variants are hosted on an ElasticSearch (Kibana v6.0) platform called Genotype to Phenotype (G2P, https://search.cancervariants.org/#*) that allows users to query and filter aggregated evidence from the various databases listed above as well as the GA4GH beacon service (http://beacon-network.org/#/), which yields access to genetic mutation data from over 200 datasets as of 2016 [@theglobalallianceforgenomicsandhealth2016]. For drug screen datasets, we focus on three well-known, large-scale drug projects: the CCLE and GDSC, which were mentioned above, and the Center for the Science of Therapeutics at the Broad Institute’s Cancer Therapeutics Response Portal (CTRP) dataset [@basu2013; @rees2016; @seashore-ludlow2015]. We also analyzed the results from a large-scale CRISPR/Cas9 screen conducted by the Broad Institute’s Cancer Dependency Map project (DepMap) using the Avana knockout library [@doench2016a; @meyers2017], which will address whether gene essentiality screens via CRISPR yield results comparable to functional drug screens.

@@

<!-- After filtering, `r nrow(g2p_filt)` level-A G2P interpretations were carried forward in subsequent analyses, constituting `r length(g2p_drugs)` unique drugs and `r length(g2p_genes)` unique genes in `r length(g2p_druggenes)` distinct level-A combinations (*Figure 1: Pink heatmap*). The number of level-A interpretations per drug ranged from `r range(g2p_drug_tally$n)[1]` to `r range(g2p_drug_tally$n)[2]` ($M =$ `r formatC(mean(g2p_drug_tally$n), digits = 2)`, $SD =$ `r formatC(sd(g2p_drug_tally$n), digits = 2)`), `r range(g2p_gene_tally$n)[1]` to `r range(g2p_gene_tally$n)[2]` ($M =$ `r formatC(mean(g2p_gene_tally$n), digits = 2)`, $SD =$ `r formatC(sd(g2p_gene_tally$n), digits = 2)`) per gene, and for each level-A drug-gene combination, the range was `r range(g2p_druggene_tally$n)[1]` to `r range(g2p_druggene_tally$n)[2]` ($M =$ `r formatC(mean(g2p_druggene_tally$n), digits = 2)`, $SD =$ `r formatC(sd(g2p_druggene_tally$n), digits = 2)`). These unique drugs, genes, and combinations are hereafter referred to as G2P drugs, G2P genes, and G2P associations. Filtering our screening datasets by these clinical criteria yields panels of potentially clinically relevant omic features. Of the `r length(g2p_drugs)` unique G2P drugs, N of them were screened in at least one of the drug screen datasets, N drugs were screened in at least two of the three datasets, and only N were screened in all three datasets. The remaining N un-screened G2P drugs suggests an incompleteness of these drug screens, though these projects frequently release new screening data that may include these drugs in future releases. -->

@@

In the `r nrow(g2p_filt)` level-A G2P interpretations, there was a wide variation in the specificity of cancer description, ranging from “Waldenström macroglobulinemia” and “hyper eosinophilic advanced syndrome” on the more detailed end to “cancer” on the broad end. Similarly, the drug screen datasets had a wide range in lineage specificity, e.g. CCLE, GTRP, and GDSC all had lineages labeled “leukemia” and “T-cell childhood acute lymphocytic leukemia” and the CRISPR dataset had cell lines labeled “Epstein-Barr virus-related Burkitt lymphoma” and “lymphoma.” To more easily conduct comparative analyses in a cancer-specific manner, we developed a more general lineage grouping method derived from the Human Disease Ontology identity codes (DOIDs, http://disease-ontology.org/) [@schriml2019] assigned to each cell line, yielding `r length(unique(ccl_converter$group_general_lineage_name))` unique lineage groupings. This custom lineage grouping was included in our harmonized cancer cell line database (see Methods).

@@

@@

## Drug screen dataset retrieval
**12/4 updated**
We utilized small molecule screening data from the Cancer Therapeutics Response Portal (CTRP, Broad Institute) [@basu_interactive_2013; @rees_correlating_2016; @seashore-ludlow_harnessing_2015], the Genomics of Drug Sensitivity in Cancer (GDSC, Wellcome Sanger Institute and Massachusetts General Hospital Cancer Center), [@garnett_systematic_2012; @yang_genomics_2012], and the Cancer Cell Line Encyclopedia (CCLE, Broad Institute and Novartis Institutes for Biomedical Research) [@barretina_cancer_2012; @consortium_pharmacogenomic_2015]. To ensure metric consistency, normalized area under the IC50 curve (AUC) for all three datasets were taken from the Quantitative Analysis of Pharmacogenomics in Cancer (http://tanlab.ucdenver.edu/QAPC/) (QAPC) [@pozdeyev_integrating_2016].

## DepMap data retrieval and processing
**12/4 updated**
CRISPR/Cas9 and mutation calls, fusions, gene expression, and copy number data from the CCLE were all retrieved from the Broad Institute’s Cancer Dependency Map (DepMap) Public 19Q2 release via the DepMap data portal (https://depmap.org/portal/download/) [@broad_depmap_depmap_2019-1]. For cell lines with no CCLE annotation, the DepMap group drew raw copy number and mutation data from whole exome sequencing data produced by the Wellcome Trust Sanger Institute [Catalogue Of Somatic Mutations In Cancer (COSMIC, https://cancer.sanger.ac.uk/cell_lines) [@bamford_cosmic_2004; @forbes_cosmic:_2017]; European Genome-phenome Archive [@lappalainen_european_2015], accession number EGAD00001001039] and processed the data following the CCLE pipelines to ensure consistency.

@@

### Cell line harmonization
**12/4 updated**
In an effort to compare CCLE, CTRP, GDSC and CRISPR results and include lineage specificity, it was necessary to consolidate cell lines screened in each dataset. Given that unique identifiers can be used between studies, it was necessary to standardize the identity of each cell line with a common source. To achieve this, a freely available cell line database called Cellosaurus (v26.0) [@bairoch_cellosaurus_2018] was downloaded and merged with the Broad Institute’s DepMap cell line data to create a harmonized cancer cell line database containing synonymous identifiers that enabled consolidation of all cell lines used in CCLE, CTRP, GDSC and CRISPR into a common framework. Subsequent modifications to this database included the merging of DepMap IDs used in the 19Q2 DepMap data release and harmonization with the consensus cell line names from @pozdeyev_integrating_2016. We curated missing synonyms and identifiers, and we manually annotated the granularity of lineage labeling for all cell lines analyzed in this study based on their Human Disease Ontology identity codes (DOIDs, http://disease-ontology.org/) [@schriml_human_2019], resulting in `r length(levels(g2p$disease))` unique lineages.

### Mutation calls
**12/4 EDIT THIS**
To preclude variation in genomic feature calls across the datasets, we used identical omics annotations from CCLE for all cell lines screened in the three datasets. The CCLE called substitutions using MuTect [@cibulskis_sensitive_2013] and annotated variants using Oncotator [@ramos_oncotator:_2015] and indels using Indelocator (https://software.broadinstitute.org/cancer/cga/indelocator).

@@

### Copy number data
**12/4 EDIT THIS**
The CCLE generated genomic copy number (CN) data using the Affymetrix Genome-Wide Human SNP Array 6.0 and GenePattern pipeline [@thecancergenomeatlasresearchnetwork2008] and normalized segmented CN log2-ratios for 23,299 genes across 1,098 cell lines using circular binary segmentation [@olshen2004] (dataset: public_18Q4_gene_cn.csv, accessed 15 November 2018).

### Gene expression data
**12/4 EDIT THIS**
The CCLE reports gene expression in RPKM (reads per kilobase per million mapped reads) for 54,356 genes across 1,156 cell lines, generated on the GeneChip Human Genome U133 Plus 2.0 Array (dataset: CCLE_DepMap_18q3_RNAseq_RPKM_20180718.gct, accessed 18 July 2018).

## G2P cancer variants
**12/4 updated**
Cancer variants with the highest level of evidence (i.e. level A) in the Genotype to Phenotype (G2P) database were retrieved from the Variant Interpretation for Cancer Consortium (VICC) portal (https://search.cancervariants.org/#*) [@wagner_harmonized_2018] using a customized JSON-query script. All JSON queries were passed to the available application program interface (API) where a request was made for all drug-gene associations with level-A evidence via the G2P evidence label, i.e. `association.evidence_label`. From manual inspection using G2P’s front-end Kibana interface, it was known that, at the time of the last query, 27 November 2018, there were 1,297 known level-A associations. Due to the limited request processing capabilities of the G2P JSON API, queries were batched into packets of 10 data points, for a total of 130 requests made in succession. Each returned request was processed as JSON object and searched to identify any existing key:value pairs for the following variables: evidence level, mutation, gene, chromosome, start, end, ref, alt, direction, phenotype description, phenotype family, phenotype ID, drug, drug ID, feature names, and sequence ID. For any evidence point where a given key:value pair was not found or unavailable, a value of -1 was assigned. One of the 1,297 level-A entries was irretrievable, yielding a final dataset of 1,296 level-A G2P interpretations, which were stored in a pandas data table and exported in CSV format. We split each interpretation by row to isolate drug associations with singular lineages, protein changes, and/or mutation types. The VICC’s methods for the harvesting and harmonizing of cancer variants is available in a GitHub repository from Oregon Health & Science University (https://github.com/ohsu-comp-bio/g2p-aggregator).

@@

## Computational
**12/4 EDIT THIS**
```{r, include=FALSE}
r_v <- paste0("v", getRversion())
# rstudio_v <- paste0("v", RStudio.Version()$version)
ggplot_v <- paste0("v", packageVersion("ggplot2"))
ggpubr_v <- paste0("v", packageVersion("ggpubr"))
```

@@

### Statistics
**12/4 EDIT THIS**
Wilcoxon tests were conducted using the `compare_means` function in the ggpubr (`r ggpubr_v`) R package [@kassambara2018]. `compare_means` does Benjamini-Hochberg p-value adjustment using the `p.adjust` function from the `stats` base R package.

>>>>

Cancer cell lines are a long-standing model for systematic testing of candidate therapeutics, beginning with the National Cancer Institute 60 (NCI60) assay from the late 1980s [@alley_feasibility_1988; @shoemaker_nci60_2006; @stinson_morphological_1992], which has been used to screen over 100,000 compounds as of 2010 [@holbeck_analysis_2010]. Since then, numerous small-molecule and gene essentiality screens of various scales and study aims have been conducted in cell lines, from grouping drugs by therapeutic target similarity [@greshock_molecular_2010] to screening only near-haploid cell lines to generate genome-level insights into gene essentiality [@wang_identification_2015] to broadly identifying cancer dependencies with large-scale screens in multiple cancer types [@mcdonald_project_2017; @mcfarland_improved_2018; @meyers_computational_2017; @patel_identification_2017] or select lineages [@heiser_integrated_2009; @marcotte_essential_2012; @patel_identification_2017]. Despite their widespread use, cancer cell lines are known to have issues with inconsistent naming conventions and contamination [@yu_resource_2015], and some cell lines have been shown to vary widely at the genetic level and in response to drug treatment across strains [@ben-david_genetic_2018]. Additionally, comparisons between cancer cell lines and tumors indicate that cell lines have higher numbers of genomic aberrations [@domcke_evaluating_2013; @mouradov_colorectal_2014; @neve_collection_2006] and tend to be hypermethylated [@paz_systematic_2003; @varley_dynamic_2013], which could prove an impediment to translating cell line discoveries into the clinic.

There have also been debates over consistency between drug screen datasets, namely the Broad Institute’s and Novartis Institutes for Biomedical Research’s Cancer Cell Line Encyclopedia (CCLE) [@barretina_cancer_2012; @consortium_pharmacogenomic_2015] and the Genomics of Drug Sensitivity in Cancer (GDSC) from the Cancer Genome Project at the Wellcome Sanger Institute and the Center for Molecular Therapeutics at Massachusetts General Hospital Cancer Center [@garnett_systematic_2012; @yang_genomics_2012]. The GDSC has also been referred to as the Cancer Genome Project (CGP) and the Sanger dataset. Studies have shown that drug-gene interactions matched between CCLE and GDSC exhibited poor correlation and inconsistencies [@haibe-kains_inconsistency_2013; @jang_systematic_2014], prompting other groups to join the debate on how best correct for experimental and methodological variation between the original drug screens and subsequent computational analysis [@consortium_pharmacogenomic_2015; @geeleher_cancer_2016; @geeleher_consistency_2016; @hatzis_enhancing_2014; @haverty_reproducible_2016; @safikhani_revisiting_2017; @safikhani_safikhani_2016]

@@

To curate an evidence-based list of these clinical biomarkers, we utilized a new oncological data aggregation project conducted by the Variant Interpretation for Cancer Consortium (VICC; https://cancervariants.org/), a Driver Project for the Global Alliance for Genomics and Health (GA4GH) [@global_alliance_for_genomics_and_health_federated_2016]. The VICC has curated annotations of known cancer variants at varying levels of evidence from multiple resources [@wagner_harmonized_2018]: the Precision Medicine Knowledgebase [@huang_cancer_2016], MolecularMatch (https://www.molecularmatch.com/index.html), OncoKB [@chakravarty_oncokb:_2017], Jackson Labs Clinical Knowledgebase [@patterson_clinical_2016], Clinical Interpretations of Variance in Cancers (CIViC) [@griffith_civic_2017], and the Cancer Genome Interpreter (CGI) [@tamborero_cancer_2018]. These harmonized variants are hosted on an ElasticSearch (Kibana v6.0) platform called Genotype to Phenotype (G2P, https://search.cancervariants.org/#*) that allows users to query and filter aggregated evidence from the various databases listed above as well as the GA4GH beacon service (http://beacon-network.org/#/), which yields access to genetic mutation data from over 200 datasets as of 2016 [@global_alliance_for_genomics_and_health_federated_2016]. For drug screen datasets, we focus on three well-known, large-scale drug projects: the CCLE and GDSC, which were mentioned above, and the Center for the Science of Therapeutics at the Broad Institute’s Cancer Therapeutics Response Portal (CTRP) dataset [@basu_interactive_2013; @rees_correlating_2016; @seashore-ludlow_harnessing_2015]. We also analyzed the results from a large-scale CRISPR/Cas9 screen conducted by the Broad Institute’s Cancer Dependency Map project (DepMap) using the Avana knockout library [@doench_optimized_2016; @meyers_computational_2017], which will address whether gene essentiality screens via CRISPR yield results comparable to functional drug screens.

@@

@@

In the `nrow(g2p_filt)` level-A G2P interpretations, there was a wide variation in the specificity of cancer description, ranging from “Waldenström macroglobulinemia” and “hyper eosinophilic advanced syndrome” on the more detailed end to “cancer” on the broad end. Similarly, the drug screen datasets had a wide range in lineage specificity, e.g. CCLE, GTRP, and GDSC all had lineages labeled “leukemia” and “T-cell childhood acute lymphocytic leukemia” and the CRISPR dataset had cell lines labeled “Epstein-Barr virus-related Burkitt lymphoma” and “lymphoma.” To more easily conduct comparative analyses in a cancer-specific manner, we developed a more general lineage grouping method derived from the Human Disease Ontology identity codes (DOIDs, http://disease-ontology.org/) [@schriml2019] assigned to each cell line, yielding ___ unique lineage groupings. This custom lineage grouping was included in our harmonized cancer cell line database (see Methods).

@@

**NUMBERS WRONG HERE**

@@

## Drug screen dataset retrieval

**12/4 updated**

We utilized small molecule screening data from the Cancer Therapeutics Response Portal (CTRP, Broad Institute) [@basu_interactive_2013; @rees_correlating_2016; @seashore-ludlow_harnessing_2015], the Genomics of Drug Sensitivity in Cancer (GDSC, Wellcome Sanger Institute and Massachusetts General Hospital Cancer Center), [@garnett_systematic_2012; @yang_genomics_2012], and the Cancer Cell Line Encyclopedia (CCLE, Broad Institute and Novartis Institutes for Biomedical Research) [@barretina_cancer_2012; @consortium_pharmacogenomic_2015]. To ensure metric consistency, normalized area under the IC50 curve (AUC) for all three datasets were taken from the Quantitative Analysis of Pharmacogenomics in Cancer (http://tanlab.ucdenver.edu/QAPC/) (QAPC) [@pozdeyev_integrating_2016].

## DepMap data retrieval and processing

**12/4 updated**

CRISPR/Cas9 and mutation calls, fusions, gene expression, and copy number data from the CCLE were all retrieved from the Broad Institute’s Cancer Dependency Map (DepMap) Public 19Q2 release via the DepMap data portal (https://depmap.org/portal/download/) [@broad_depmap_depmap_2019-1]. For cell lines with no CCLE annotation, the DepMap group drew raw copy number and mutation data from whole exome sequencing data produced by the Wellcome Trust Sanger Institute [Catalogue Of Somatic Mutations In Cancer (COSMIC, https://cancer.sanger.ac.uk/cell_lines) [@bamford_cosmic_2004; @forbes_cosmic:_2017]; European Genome-phenome Archive [@lappalainen_european_2015], accession number EGAD00001001039] and processed the data following the CCLE pipelines to ensure consistency.

@@

### Cell line harmonization

**12/4 updated**

In an effort to compare CCLE, CTRP, GDSC and CRISPR results and include lineage specificity, it was necessary to consolidate cell lines screened in each dataset. Given that unique identifiers can be used between studies, it was necessary to standardize the identity of each cell line with a common source. To achieve this, a freely available cell line database called Cellosaurus (v26.0) [@bairoch_cellosaurus_2018] was downloaded and merged with the Broad Institute’s DepMap cell line data to create a harmonized cancer cell line database containing synonymous identifiers that enabled consolidation of all cell lines used in CCLE, CTRP, GDSC and CRISPR into a common framework. Subsequent modifications to this database included the merging of DepMap IDs used in the 19Q2 DepMap data release and harmonization with the consensus cell line names from @pozdeyev_integrating_2016. We curated missing synonyms and identifiers, and we manually annotated the granularity of lineage labeling for all cell lines analyzed in this study based on their Human Disease Ontology identity codes (DOIDs, http://disease-ontology.org/) [@schriml_human_2019], resulting in `r length(levels(g2p$disease))` unique lineages.

### Mutation calls

**12/4 EDIT THIS**

To preclude variation in genomic feature calls across the datasets, we used identical omics annotations from CCLE for all cell lines screened in the three datasets. The CCLE called substitutions using MuTect [@cibulskis_sensitive_2013] and annotated variants using Oncotator [@ramos_oncotator:_2015] and indels using Indelocator (https://software.broadinstitute.org/cancer/cga/indelocator).

@@

### Copy number data

**12/4 EDIT THIS**

The CCLE generated genomic copy number (CN) data using the Affymetrix Genome-Wide Human SNP Array 6.0 and GenePattern pipeline [@thecancergenomeatlasresearchnetwork2008] and normalized segmented CN log2-ratios for 23,299 genes across 1,098 cell lines using circular binary segmentation [@olshen2004] (dataset: public_18Q4_gene_cn.csv, accessed 15 November 2018).

### Gene expression data

**12/4 EDIT THIS**

The CCLE reports gene expression in RPKM (reads per kilobase per million mapped reads) for 54,356 genes across 1,156 cell lines, generated on the GeneChip Human Genome U133 Plus 2.0 Array (dataset: CCLE_DepMap_18q3_RNAseq_RPKM_20180718.gct, accessed 18 July 2018).

## G2P cancer variants

**12/4 updated**

Cancer variants with the highest level of evidence (i.e. level A) in the Genotype to Phenotype (G2P) database were retrieved from the Variant Interpretation for Cancer Consortium (VICC) portal (https://search.cancervariants.org/#*) [@wagner_harmonized_2018] using a customized JSON-query script. All JSON queries were passed to the available application program interface (API) where a request was made for all drug-gene associations with level-A evidence via the G2P evidence label, i.e. `association.evidence_label`. From manual inspection using G2P’s front-end Kibana interface, it was known that, at the time of the last query, 27 November 2018, there were 1,297 known level-A associations. Due to the limited request processing capabilities of the G2P JSON API, queries were batched into packets of 10 data points, for a total of 130 requests made in succession. Each returned request was processed as JSON object and searched to identify any existing key:value pairs for the following variables: evidence level, mutation, gene, chromosome, start, end, ref, alt, direction, phenotype description, phenotype family, phenotype ID, drug, drug ID, feature names, and sequence ID. For any evidence point where a given key:value pair was not found or unavailable, a value of -1 was assigned. One of the 1,297 level-A entries was irretrievable, yielding a final dataset of 1,296 level-A G2P interpretations, which were stored in a pandas data table and exported in CSV format. We split each interpretation by row to isolate drug associations with singular lineages, protein changes, and/or mutation types. The VICC’s methods for the harvesting and harmonizing of cancer variants is available in a GitHub repository from Oregon Health & Science University (https://github.com/ohsu-comp-bio/g2p-aggregator).

@@

## Computational

**12/4 EDIT THIS**
```{r, include=FALSE}
r_v <- paste0("v", getRversion())
# rstudio_v <- paste0("v", RStudio.Version()$version)
ggplot_v <- paste0("v", packageVersion("ggplot2"))
ggpubr_v <- paste0("v", packageVersion("ggpubr"))
```

@@

### Statistics

**12/4 EDIT THIS**

Wilcoxon tests were conducted using the `compare_means` function in the ggpubr (`r ggpubr_v`) R package [@kassambara2018]. `compare_means` does Benjamini-Hochberg p-value adjustment using the `p.adjust` function from the `stats` base R package.
